---
title: "hw02_explore-gapminder-dplyr"
author: "Gokul Raj Suresh Kumar"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Explore Gapminder with dplyr

## Bringing rectangular data in

#### Loading the gapminder and dplyr packages 

```{r}
suppressPackageStartupMessages(library(dplyr))
library(gapminder)
```

## Smell testing the data

#### Displaying a tibble
```{r}
gapminder
```
- A `data.frame` is a list of vectors of equal length. 
- Printing the `gapminder` object prints a tibble. 
- A tibble is a `data.frame` reimagined to retrieve just the effective information.

#### Determining the class 

```{r}
class(gapminder)
```
- The `gapminder` object is a regular data frame.
- It is a tibble as well.

#### Determining the number of variables

```{r}
ncol(gapminder)
```

#### Determining the number of observations

```{r}
nrow(gapminder)
```

#### Alternate ways of determining the size

```{r}
dim(gapminder)
```

- The 2 functions described in the subsequent section can also be considered as alternate ways of determining the size.

#### Determining the flavour of each variable

```{r}
str(gapminder)
glimpse(gapminder)
```
1. *country* is of type *Factor*.
2. *continent* is of type *Factor*.
3. *year* is of type *int*.
4. *lifeExp* is of type *num*.
5. *pop* is of type *int*.

## Exploring individual variables
