---
title: "hw02_explore-gapminder-dplyr"
author: "Gokul Raj Suresh Kumar"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Explore Gapminder with dplyr

## Bringing rectangular data in

#### Loading the gapminder and dplyr packages 

```{r}
suppressPackageStartupMessages(library(dplyr))
library(gapminder)
```

## Smell testing the data

#### Displaying a tibble
```{r}
gapminder
```
- A `data.frame` is a list of vectors of equal length. 
- Printing the `gapminder` object prints a tibble. 
- A tibble is a `data.frame` reimagined to retrieve just the effective information.

```{r}
summary(gapminder)
```


#### Determining the class 

```{r}
class(gapminder)
```
- The `gapminder` object is a regular data frame.
- It is a tibble as well.

#### Determining the number of variables

```{r}
ncol(gapminder)
```

#### Determining the number of observations

```{r}
nrow(gapminder)
```

#### Alternate ways of determining the size

```{r}
dim(gapminder)
```

- The 2 functions described in the subsequent section can also be considered as alternate ways of determining the size.

#### Determining the flavour of each variable

```{r}
str(gapminder)
glimpse(gapminder)
```
1. **country** is of type **Factor**.
2. **continent** is of type **Factor**.
3. **year** is of type **int**.
4. **lifeExp** is of type **num**.
5. **pop** is of type **int**.

## Exploring individual variables

The variables **pop** and **lifeExp** have been chosen as quantitative variables for this analysis.

```{r}
class(gapminder$pop)
summary(gapminder$pop)
quantile(gapminder$pop)
mean(gapminder$pop)
median(gapminder$pop)
class(gapminder$lifeExp)
summary(gapminder$lifeExp)
quantile(gapminder$lifeExp)
mean(gapminder$lifeExp)
median(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

The variables **continent** and **country**  have been chosen as categorical variables for this analysis.

```{r}
summary(gapminder$continent)
class(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$country)
barplot(table(gapminder$continent))
class(gapminder$country)
nlevels(gapminder$country)
```


## Exploring various plot types

#### Scatterplots
```{r}
p <- ggplot(filter(gapminder,continent == "Asia"), aes(x = gdpPercap , y = lifeExp))
p + geom_point(alpha = 1/4, size = 4) + scale_x_log10() + 
  geom_smooth(lwd = 2, se = FALSE)
```

#### 




