---
title: "Tidy data and joins"
author: "Gokul Raj Suresh Kumar"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Tidying data and performing joins

####Loading the tidyverse and gapminder packages

```{r message=FALSE}
library( tidyverse )
library( gapminder )
```

##General data reshaping and relationship to aggregation

###Activity #2 

####Life expectancy for different countries by year

```{r}
different_countries <- gapminder %>% 
  select( year , country , lifeExp ) %>% 
  filter( country %in% c( "India" , "Canada" , "United States" , "China" , "Japan" ) )

le_by_year <- different_countries %>% 
  spread ( key = "country" , value = "lifeExp" ) %>% 
  rename( Year = year )

knitr::kable( le_by_year )

le_by_year %>% ggplot( aes( x = Year , y = India , color = "India" ) ) +
  geom_point( ) + geom_line( ) +
  geom_point( aes( x = Year , y = Canada , color = "Canada" ) ) +  
  geom_line( aes( x = Year , y = Canada , color = "Canada" ) ) +
  scale_color_manual( name = "" , values = c( "India" = "red" ,
                                              "Canada" = "blue" ) ) +
  labs( title = "Life Expectancy over Time" , x = "Year" , y = "lifeExp" )
```

###Activity #3

####Mean life expectancy for all years by continent

```{r}
all_continents <- gapminder %>% 
  select( continent , year , lifeExp ) %>% 
  group_by( year , continent ) %>% 
  summarise( mean_lifeExp = mean( lifeExp ) ) %>% 
  arrange( continent )

le_continents <- all_continents %>% 
  spread( key = "continent" , value = "mean_lifeExp") %>% 
  rename( Year = year )

knitr::kable( le_continents )

le_continents %>% ggplot( aes( x = Year , y = Africa , color = "Africa") ) +
  geom_point( ) + geom_line( ) +
  geom_point( aes( x = Year , y = Americas , color = "Americas" ) ) +
  geom_line( aes( x = Year , y = Americas , color = "Americas" ) ) +
  geom_point( aes( x = Year , y = Asia , color = "Asia" ) ) +
  geom_line( aes( x = Year , y = Asia , color = "Asia" ) ) +
  geom_point( aes( x = Year , y = Europe , color = "Europe" ) ) +
  geom_line( aes( x = Year , y = Europe , color = "Europe" ) ) +
  geom_point( aes( x = Year , y = Oceania , color = "Oceania" ) ) +
  geom_line( aes( x = Year , y = Oceania , color = "Oceania" ) ) +
  scale_color_manual( name = "" , values = c( "Africa" = "red" , 
                                              "Americas" = "green" ,
                                              "Asia" = "orange" , 
                                              "Europe" = "blue" ,
                                              "Oceania" = "brown") ) + 
    labs( title = "Life Expectancy over Time" , x = "Year" , y = "Life Expectancy" )
```

###Activity #4

#### Continents having lowest and highest life expectancy by year

```{r}
min_max_le <- gapminder %>%
  select( year , continent , lifeExp ) %>%
  group_by( year ) %>%
  filter( min_rank( desc( lifeExp) ) == 1 | min_rank( lifeExp ) == 1 ) %>% 
  arrange( year )

min_max_le_continent <- min_max_le %>% 
  spread( key = "continent" , value = "lifeExp" ) %>% 
  rename( Year = year )

knitr::kable( min_max_le_continent )
```

##Join, merge, look up

###Activity #1

####Joining Gapminder with a complimentary data frame

```{r}
knitr::kable( gapminder_data <- gapminder %>% 
  filter( continent == "Asia" , year == 2007 , lifeExp > 75 ) %>% 
  select( country , lifeExp, gdpPercap ) )

country_info <- "
     country, capitol_city, language_spoken
     Bahrain,       Manama,          Arabic
       Japan,        Tokyo,        Japanese
      Israel,    Jerusalem,          Hebrew
      Taiwan,       Taipei,        Mandarin
       India,    New Delhi,           Hindi
    Cambodia,   Phnom Penh,           Khmer
    Pakistan,    Islamabad,            Urdu
"
country_info <- read_csv( country_info , trim_ws = TRUE , skip = 1 )

knitr::kable( country_info )

knitr::kable( inner_join( gapminder_data , country_info ) )

knitr::kable( semi_join( gapminder_data , country_info ) )

knitr::kable( left_join( gapminder_data , country_info ) )

knitr::kable( anti_join( gapminder_data , country_info ) )

knitr::kable( inner_join( country_info , gapminder_data ) )

knitr::kable( semi_join( country_info , gapminder_data ) )

knitr::kable( left_join( country_info , gapminder_data ) )

knitr::kable( anti_join( country_info , gapminder_data ) )

knitr::kable( full_join( country_info , gapminder_data ) )
```


