---
title: "Tidy data and joins"
author: "Gokul Raj Suresh Kumar"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Tidying data and performing joins

####Loading the tidyverse and gapminder packages

```{r message=FALSE}
library( tidyverse )
library( gapminder )
```

##General data reshaping and relationship to aggregation

###Activity #2 

####Life expectancy for different countries by year

```{r dev='CairoPNG'}
# filtering down to 5 countries to make life easier
different_countries <- gapminder %>% 
  select( year , country , lifeExp ) %>% 
  filter( country %in% c( "India" , "Canada" , "United States" , "China" , "Japan" ) ) %>% 
  print( n = 10 )
# making a tibble with one row per year and columns for life expectancy for 5 countries
le_by_year <- different_countries %>% 
  spread ( key = "country" , value = "lifeExp" ) %>% 
  rename( Year = year )
knitr::kable( le_by_year )
# plotting life expectancy for India against that for Canada
le_by_year %>% ggplot( aes( x = Year , y = India , color = "India" ) ) +
  geom_point( ) + geom_line( ) +
  geom_point( aes( x = Year , y = Canada , color = "Canada" ) ) +  
  geom_line( aes( x = Year , y = Canada , color = "Canada" ) ) +
  scale_color_manual( name = "" , values = c( "India" = "red" ,
                                              "Canada" = "blue" ) ) +
  labs( title = "Life Expectancy over Time" , x = "Year" , y = "lifeExp" )
```

###Activity #3

####Mean life expectancy for all years by continent

```{r dev='CairoPNG'}
# mean life expectancy for each continent by year
all_continents <- gapminder %>% 
  select( continent , year , lifeExp ) %>% 
  group_by( year , continent ) %>% 
  summarise( mean_lifeExp = mean( lifeExp ) ) %>% 
  arrange( continent ) %>% 
  print( n = 10 )
# reshaping to one row per year and one variable for each continent
le_continents <- all_continents %>% 
  spread( key = "continent" , value = "mean_lifeExp") %>% 
  rename( Year = year )
knitr::kable( le_continents )
# trying to plot the life expectancy over time for different continents !
# this attempt is definitely not the most efficient way to go about this ! 
# is there a better way ? I wonder . .
le_continents %>% ggplot( aes( x = Year , y = Africa , color = "Africa") ) +
  geom_point( ) + geom_line( ) +
  geom_point( aes( x = Year , y = Americas , color = "Americas" ) ) +
  geom_line( aes( x = Year , y = Americas , color = "Americas" ) ) +
  geom_point( aes( x = Year , y = Asia , color = "Asia" ) ) +
  geom_line( aes( x = Year , y = Asia , color = "Asia" ) ) +
  geom_point( aes( x = Year , y = Europe , color = "Europe" ) ) +
  geom_line( aes( x = Year , y = Europe , color = "Europe" ) ) +
  geom_point( aes( x = Year , y = Oceania , color = "Oceania" ) ) +
  geom_line( aes( x = Year , y = Oceania , color = "Oceania" ) ) +
  scale_color_manual( name = "" , values = c( "Africa" = "red" , 
                                              "Americas" = "green" ,
                                              "Asia" = "orange" , 
                                              "Europe" = "blue" ,
                                              "Oceania" = "brown") ) + 
    labs( title = "Life Expectancy over Time" , x = "Year" , y = "Life Expectancy" )
```

**Take home message**:

**Listen to Jenny**. When she said that **tidy data is integral to efficient data analysis and visualization**, she meant it.

###Activity #4

#### Continents having lowest and highest life expectancy by year

```{r}
# the tibble referred-to by Jenny, modified to give continent with 
# lowest and highest life expectancy by year
min_max_le <- gapminder %>%
  select( year , continent , lifeExp ) %>%
  group_by( year ) %>%
  filter( min_rank( desc( lifeExp) ) == 1 | min_rank( lifeExp ) == 1 ) %>% 
  arrange( year ) %>% 
  print( n = 10 )
# reshaping the above tibble to get one row per year 
min_max_le_continent <- min_max_le %>% 
  spread( key = "continent" , value = "lifeExp" ) %>% 
  rename( Year = year )
knitr::kable( min_max_le_continent )
```

`NA`'s are returned for countries that don't have the minimum or maximum life expectancy for a specific year.

##Join, merge, look up

###Activity #1

####Joining Gapminder with a complimentary data frame

```{r}
# filtering Asian countries which had life expectancy greater than 75 in 2007
knitr::kable( gapminder_data <- gapminder %>% 
  filter( continent == "Asia" , year == 2007 , lifeExp > 75 ) %>% 
  select( country , lifeExp, gdpPercap ) )
# creating the complementary dataframe for gapminder
country_info <- "
     country, capitol_city, language_spoken
     Bahrain,       Manama,          Arabic
       Japan,        Tokyo,        Japanese
      Israel,    Jerusalem,          Hebrew
      Taiwan,       Taipei,        Mandarin
       India,    New Delhi,           Hindi
    Cambodia,   Phnom Penh,           Khmer
    Pakistan,    Islamabad,            Urdu
"
country_info <- read_csv( country_info , trim_ws = TRUE , skip = 1 )
knitr::kable( country_info )
```

```{r}
# inner_join( gapminder_data , country_info )
knitr::kable( inner_join( gapminder_data , country_info ) )
```

We lose `Hong Kong`, `Korea`, `Kuwait`, `Oman` and `Singapore` because their names don't appear in `country_info`. The result has all variables from `gapminder_data` plus `capitol_city` and `language_spoken` from `country_info`.

```{r}
# semi_join( gapminder_data , country_info )
knitr::kable( semi_join( gapminder_data , country_info ) )
```

Results are similar to the above `inner_join`'s result, but variables from`country_info` are not returned.

```{r}
# left_join( gapminder_data , country_info )
knitr::kable( left_join( gapminder_data , country_info ) )
```

`gapminder_data` is returned along with `capitol_city` and `language_spoken` from `country_info` with `NA`'s for countries that don't appear in `country_info`.

```{r}
# anti_join( gapminder_data , country_info )
knitr::kable( anti_join( gapminder_data , country_info ) )
```

Countries from `gapminder_data` that don't appear in `country_info` are returned, without `capitol_city` and `language_spoken`.

```{r}
# inner_join( country_info , gapminder_data )
knitr::kable( inner_join( country_info , gapminder_data ) )
```

We lose `Hong Kong`, `Korea`, `Kuwait`, `Oman` and `Singapore` because their names don't appear in `country_info`. The result has all variables from `country_info` plus `lifeExp` and `gdpPercap` from `gapminder_data`.

```{r}
# semi_join( country_info , gapminder_data )
knitr::kable( semi_join( country_info , gapminder_data ) )
```

Results are similar to the above `inner_join`'s result, but variables from`gapminder_data` are not returned.

```{r}
# left_join( country_info , gapminder_data )
knitr::kable( left_join( country_info , gapminder_data ) )
```

`country_info` is returned along with `lifeExp` and `gdpPercap` from `gapminder_data` with `NA`'s for countries that don't appear in `gapminder_data`.


```{r}
# anti_join( country_info , gapminder_data )
knitr::kable( anti_join( country_info , gapminder_data ) )
```

Countries from `country_info` that don't appear in `gapminder_data` are returned, without `lifeExp` and `gdpPercap`.

```{r}
# full_join( gapminder_data , country_info )
knitr::kable( full_join( gapminder_data , country_info ) )
```

We get all rows of `gapminder_data` and 3 rows from `country_info` (`India`, `Cambodia` and `Pakistan`) with columns `capitol_city` and `language_spoken`. `NA`'s are returned for rows that derive solely from either tables.

##Reflections

I got super curious and ended up solving multiple prompts. Apologies if my lengthy document having so many activities annoys the reviewer. Knowledge gained through this assignment was huge! 

Jenny's [ggplot2 tutorial](https://github.com/jennybc/ggplot2-tutorial) and her [Cheatsheet for dplyr join functions](http://stat545.com/bit001_dplyr-cheatsheet.html) were of massive help. I wonder how the figures in her tutorial look so polished ! I tried ! 

Update[18 Oct 2016]: I was able to enable aliasing on windows using `Cairo` to get nice plots.

I referred to [Beautiful plotting in R: A ggplot2 cheatsheet](http://www.cs.utexas.edu/~cannata/dataVis/Class%20Notes/Beautiful%20plotting%20in%20R_%20A%20ggplot2%20cheatsheet%20_%20Technical%20Tidbits%20From%20Spatial%20Analysis%20&%20Data%20Science.pdf) to manually set colors in plots. 
