---
title: "hw02_explore-gapminder-dplyr"
author: "Gokul Raj Suresh Kumar"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Explore Gapminder with dplyr

##Bringing rectangular data in

####Loading the gapminder and dplyr packages 

```{r}
suppressPackageStartupMessages(library(dplyr))
library(gapminder)
```

##Smell testing the data

####Displaying a tibble
```{r}
gapminder
```
- A `data.frame` is a list of vectors of equal length. 
- Printing the `gapminder` object prints a tibble. 
- A tibble is a `data.frame` reimagined to retrieve just the effective information.

```{r}
summary(gapminder)
```


####Determining the class 

```{r}
class(gapminder)
```
- The `gapminder` object is a regular data frame.
- It is a tibble as well.

####Determining the number of variables

```{r}
ncol(gapminder)
```

####Determining the number of observations

```{r}
nrow(gapminder)
```

####Alternate ways of determining the size

```{r}
dim(gapminder)
```

- The 2 functions described in the subsequent section can also be considered as alternate ways of determining the size.

####Determining the flavour of each variable

```{r}
str(gapminder)
glimpse(gapminder)
```
1. **country** is of type **Factor**.
2. **continent** is of type **Factor**.
3. **year** is of type **int**.
4. **lifeExp** is of type **num**.
5. **pop** is of type **int**.

##Exploring individual variables

The variables **pop** and **lifeExp** have been chosen as quantitative variables for this analysis.

```{r}
class(gapminder$pop)
summary(gapminder$pop)
quantile(gapminder$pop)
mean(gapminder$pop)
median(gapminder$pop)
class(gapminder$lifeExp)
summary(gapminder$lifeExp)
quantile(gapminder$lifeExp)
mean(gapminder$lifeExp)
median(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

The variables **continent** and **country**  have been chosen as categorical variables for this analysis.

```{r dev='CairoPNG'}
summary(gapminder$continent)
class(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$country)
barplot(table(gapminder$continent))
class(gapminder$country)
nlevels(gapminder$country)
```


##Exploring various plot types

####Loading `ggplot2`

```{r}
library(ggplot2)
```

####Scatterplots

```{r dev='CairoPNG'}
p <- ggplot(filter(gapminder,continent == "Asia"), aes(x = gdpPercap , y = lifeExp))
p + geom_point(alpha = 1/4, size = 4) + scale_x_log10() + geom_smooth(lwd = 2, se = FALSE)
```

####Stripplots

```{r dev='CairoPNG'}
p <- ggplot(gapminder, aes(x = continent, y = pop)) 
p + geom_jitter()
```

####Boxplots

```{r dev='CairoPNG'}
p <- ggplot(filter(gapminder, continent != "Asia"), aes(x = continent, y = lifeExp)) 
p + geom_boxplot()
```

####Histogram

```{r dev='CairoPNG'}
p <- ggplot(gapminder , aes(x = pop)) 
p + geom_histogram() + scale_x_log10()
```

##Using `filter()` , `select()` and `%>%`

####`filter()`

```{r}
filter(gapminder, country %in% c("India" , "Canada" ), year == 2007)
```

####`select()`

```{r}
select(gapminder, country , pop)
```

####`filter()` and `select()` 

```{r}
filter(select(gapminder, country, year, pop), country == "India" , year == 2007)
select(filter(gapminder, country == "India" , year == 2007), country , year , pop)
```

####The **PIPE** `%>%`

```{r}
gapminder %>% filter(country == "India" , year == 2007)
gapminder %>% select(country , pop)
```

####`filter()` , `select()` and `%>%`  

```{r}
gapminder %>% select(country , year , pop) %>% filter(country == "India", year == 2007)
gapminder %>% filter(country == "India" , year == 2007) %>% select( country , year, pop)
```

####Piping into `ggplot()`

```R
p <- ggplot(filter(gapminder, continent != "Asia"), aes(x = continent, y = lifeExp)) 
p + geom_boxplot()
```
Implementing the above boxplot I had done earlier, using the pipe.

```{r dev='CairoPNG'}
p <- gapminder %>% filter(continent != "Asia") %>% ggplot(aes(x = continent, y = lifeExp))
p + geom_boxplot()
```

##Venturing a bit further

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))

```

I believe that the Analyst's script is not fetching all the data that she needs. In order to fetch the complete data after applying the constraints, she has to replace the `==` with the `%in%`.

```{r}
filter(gapminder, country %in% c("Rwanda" , "Afghanistan"))
```

## The Process

I have fallen in love with the `%>%` ! Oddly satisfying to use it in my scripts ! This assignment did involve a decent amount of exploration, and I didn't find any technical hurdles. The resources provided by the STAT 545 team was more than sufficient for me to finish this one.



